\textls[-0]{\thetitle}

### 2023-03-31 lang-design.tex
Add back partial recursor

%\begin{centered}
%\begin{minipage}{.97\textwidth}
%\newcommand\hlcase[1]{\tikzmarkin[disable rounded corners=true,set fill color=yellow!18,set border color=yellow!18]{#1}(0.00,-0.05)(-0.00,0.20)}
%\begin{lstlisting}[basicstyle=\fontsize{8.25}{9.9}\ttfamily]
%tm_prect_STLC : \forall (P : tm -> Type), [@\hlcase{tm-partialrec-case-unit}@]option (P tm_unit)[@\tikzmarkend{tm-partialrec-case-unit}@] -> [@\hlcase{tm-partialrec-case-var}@](\forall x, option (P (tm_var x)))[@\tikzmarkend{tm-partialrec-case-var}@] ->
%                [@\hlcase{tm-partialrec-case-abs}@](\forall x t, option (P t) -> option (P (tm_abs x t)))[@\tikzmarkend{tm-partialrec-case-abs}@] ->
%                [@\hlcase{tm-partialrec-case-app}@](\forall t1, option (P t1) -> \forall t2, option (P t2) -> option (P (tm_app t1 t2)))[@\tikzmarkend{tm-partialrec-case-app}@] ->
%                \forall t, option (P t)
%\end{lstlisting}
%\end{minipage}
%\end{centered}

>>>>

\noindentparagraph{Property about Partial Recursors.}

Partial recursor is simply adding partiality into the signature of the eliminator, by wrapping each motive with "option". This partial recursor seems non-canonical compared to the (dependent) eliminator for Inductive type as the latter can be considered as universal property. This extensionality makes dependent eliminator one of \textit{the most powerful} reasoning tools around inductive type.

Even though we cannot prove partial recursors are \textit{the
most powerful} reasoning tool for extensible inductive types,
the fact that propositional partial
recursors can derive injectivity alreadt demonstrate itself as a good \textit{extensional characterization} of
(non-indexed) extensible inductive type (under certain
constraints)---because we can ``embed'' vanilla inductive types into
types supporting a partial recursor: a bit more formally, and
restricting our focus to non-indexed inductive types:

\begin{theorem}\label{thm:prec-complete} Given a list of $n$ pairs of types $\{ x : "A"_i \vdash "B"_i(x) \}_{i}$ s.t.

  \begin{itemize}
    \item \textlabel{(Detectable Partiality)}{prop:detectable-partiality} for each pair $x : A_i \vdash B_i(x)$,
    \begin{minted}{Coq}
      Axiom detectable:
      forall {T} {a} (f: Bᵢ a -> option T),
        {forall x, f x <> None} + {exists x, f x = None}.
    \end{minted} 
    \item Define $"C"$ as the inductive type using this list, i.e.,
    \begin{minted}{Coq}
Inductive C : Set := ... | cᵢ : forall (x : Aᵢ), (Bᵢ x -> C) -> C | ...
    \end{minted}
    \item Assume an arbitrary type \mintinline{Coq}{D : Set} with
    \begin{itemize}
      \item $n$ \textbf{functions} \mintinline{Coq}{dᵢ : forall (x : Aᵢ), (Bᵢ x -> D) -> D}  
      \item a partial recursor
      \begin{minted}[escapeinside=@@]{Coq}
prec : forall (R : Set), ..,
  (rᵢ : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)), ..,@\YZ{Can R be indexed?}\EDJreply{This part actually currently incorrect and needs an extra constraint. Let me fix other parts first}\YZreply{Is this constraint Detectable Partiality?}@
      D -> option R
      \end{minted}
      \item and $n$ (propositional) computational axioms: for all $i$, 
      \begin{minted}{Coq}
prec {R} r₁ r₂ .. rₙ (dᵢ aᵢ bᵢ) 
    = rᵢ aᵢ (fun x => prec {R} r₁ r₂ .. rₙ (bᵢ x))
      \end{minted}
    % where \mintinline{Coq}{(lift rᵢ) : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)} is defined as expected
    \end{itemize}
  \end{itemize}
  Then there exists an \textbf{embedding} from "C" to "D".
  More concretely, we can have \mintinline{Coq}{inj : C -> D} and 
  \mintinline{Coq}{linv : D -> option C} defined using the eliminator of\/
  "C" and "prec" of\/ "D" (both acting like identity) such that
  \begin{minted}[escapeinside=@@]{Coq}
    forall c : C, linv (inj c) = some c@\YZ{Does 'linv' correpsond to 'reflect' above? If so, then the reader should be explicitly reminded of this fact rather than having to guess it.}\EDJreply{Added below. I mentioned left inverse 'linv' is the generalization}@
  \end{minted}
\end{theorem}
The reason we have this weird looking \ref{prop:detectable-partiality}
is due to the construction of "linv"---at one point, using "prec" to
construct "linv", we need to prove \mintinline{Coq}{∀ a, (Bᵢ a → option
C) → option C}, generally not provable, unless we have
\mintinline{Coq}{option (Bᵢ a → C)} instead.\YZ{Don't you really mean
"unless we have the Detectable Partiality assumption instead"?}
This \ref{prop:detectable-partiality} is exactly transforming the former to
the latter:
recall that, according to the semantic of inductive type,
\mintinline{Coq}{(Bᵢ a → option C)} is actually the result of using
"linv" on the substructure of "D". This function points out each of the
translation result from "D" to "C" and some of them are "None" because
of failing. For us, we only care about if \textbf{any} of the
substructure of "D" fails to translate---if any substructure fails,
then the whole translation fails and thus we should have "None" as
return.

\ref{prop:detectable-partiality} is derivable for all the finitary branching inductive type\footnote{By simply enumerating through the finite domain "Bᵢ a"}---that includes all of the examples shown in this paper.

\ref{prop:detectable-partiality} is not generally supported when using function (infinitary branching) in the constructors, for example, the case of using \textit{higher order abstract syntax} to represent functions in "tm". However, in those cases, we still have partial recursor provable and useful---we just cannot see the partial recursor as a strong toolkit anymore.

We emphasize that in \cref{thm:prec-complete}, each function $"d"_i$ can
actually be considered as a constructor, because they can be reflected
to real constructors of "C" using the left inverse "linv". Thus, \cref{thm:prec-complete}
implies that (1) every future extension of the inductive type can
support this partial recursion (trivially); (2) every type supporting
this partial recursor with its computational axioms at least supports
these constructors because of the embedding. In other words, \textit{a
type (at least) supports these constructors if and only if this type
supports the corresponding partial recursor (with the computational
axioms)}.  Thus, we can argue that the partial recursor gives a sound and
complete extensional characterization of all the extension of a given (non-indexed)
inductive type.\YZ{What about indexed inductive types?}\EDJreply{Done. I also have a formalization of this proof now, for indexed type and dependent eliminator. Though the formalization is about single constructor.}

We can support partial dependent eliminator as well, but partial dependent eliminator can deduce "prec" and our "prec" is enough to contruct this left inverse for (non-indexed) inductive type. 
\begin{minted}{Coq}
  pdelim : ∀ (P : D -> Set), .., 
  (rᵢ : (∀ (x : Aᵢ) (w : Bᵢ x -> D), (∀ (b : Bᵢ x), option (P (w b))) 
    -> option (P (dᵢ x w)))),  .., ∀ (d : D) -> option (P d)
\end{minted}
To generalize to indexed inductive type, we need to base on indexed W type~\cite{martin1982constructive, morris2009indexed,jashug2017} and dependent eliminator is unavoidable. 
% Please refer to the supplementary material and appendix for a formalization of the left inverse in the context of indexed inductive type.
